---
title: "P8106 Midterm Project: Predicting COVID-19 Recovery Time"
author: 
  - "Guadalupe Antonio Lopez, Gustavo Garcia-Franceschini, Derek Lamb"
  -  "UNI's: GA2612, GEG2145, DRL2168"
header-includes:
    - \usepackage{setspace}\doublespacing

output: pdf_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(caret)
library(rsample)
library(corrplot)
library(gtsummary)

knitr::opts_chunk$set(
  echo = TRUE, 
  fig.width = 6,
  out.width = "65%")
```

## Introduction

## EDA

We also examined the pairwise correlations of the variables, and the correlations of the covariates with the recovery time. There were two clusters of strong correlation (height, weight, and BMI; hypertension and SBP), but these covariates were functionally dependent upon each other. There were no other strong correlations between variables, and no one covariate had an exceptional correlation to recovery time.

```{r corrplot, echo=FALSE}
# convert covariates to numeric
cor_rec <- df_rec |> 
  model.matrix(recovery_time ~ ., data = _) 

# put outcome back into matrix
cor_rec[,1] = df_rec$recovery_time
colnames(cor_rec)[1] <- "recovery"

# create corrplot
cor_rec |> 
  cor() |> 
  corrplot()
```

## Model Training

## Results 

Remember to talk about `study` as a variable.

## Conclusion
